### 2018-07-09 chapters/01-introduction/intro.tex
Apply article format and create a clean version

A typical workflow of a scientist who writes a scientific publication is to use a \gls{wp} to write the publication and one or more \gls{cas} for Verification, Analysis and Visualization, among other tasks. his formulae. Especially in \gls{stem}, \LaTeX\footnote{Note that technically \LaTeX{} is not a \gls{wp} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the standard to write in \gls{stem} and we only focus on mathematical writings in this thesis, we categorize \LaTeX{} as a \gls{wp} as well.} has become the de facto standard for writing scientific publications over the last 30 years~\cites{LATEX:Standard}[559]{DigitalTypo}{Knuth}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider a Jacobi polynomial~\cite[(18.3), table 1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)).
\end{equation}
This formula is written in \LaTeX{} as
\begin{equation}\label{eq:P-tex}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation}

While \LaTeX{} focuses on displaying mathematics, \gls{cas} focus on computations and user friendly syntax. Therefore, each system (such as \LaTeX{}, \Maple{} or \Mathematica) uses its own representation and syntax. Hence, a writer needs to constantly translate mathematical expressions from one representation to another and back again. Table~\ref{tab:JacobiP-usecase} shows four different representations for the same formula~(\ref{eq:P}).

@@

Such translations can become complex, if there are non-obvious differences between the used systems. For example, the \gls{cas} could use a different order of the arguments, define a normalization of the function for better performances or even use different domains for the variables. Multivalued functions are particularly difficult~\cite{AISC:MultivaluedFunctions}. A \gls{cas} usually defines so called \textit{branch cuts} to compute principal values of multivalued functions~\cite{Maple:Cuts}. Thereby, implementations of multivalued functions are discontinuous. In general, positioning branch cuts follow some conventions. They are mostly defined conveniently, such as a straight line. However, since a branch cut is a curve, it also can be defined as a spiral or it can be defined for a different position in the complex plane. The position of branch cuts can therefore varies from \gls{cas} to \gls{cas}~\cite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{} defines the branch cut at ${[-\infty\iunit, -\iunit]}$, ${[\iunit,\infty\iunit]}$ (figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (figure~\ref{fig:acot-cut2}).

@@

Hence, a \gls{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls{cas} to avoid mistakes during a translation~\cite{Maple:Cuts}. In consequence, a manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named to automatic \gls{p2c} conversion~\cite{POM-Tagger}.

An automatic translation process is desirable. Providing translations for mathematical \LaTeX{} expressions is difficult, because the semantic information is absent. Semantics is the meaning of an expression. Consider the example of the Jacobi polynomial in~(\ref{eq:P}). A scientific reader might be able to conclude information like $P$ indicating the Jacobi polynomial and $n$ is being a non-negative integer. However, for a computer it is difficult to understand the expression~(\ref{eq:P-tex}). Since \LaTeX{} is the de facto standard for mathematical expressions, most \gls{cas} provide import and export functions for \LaTeX{} expressions~\cite{Maple:ImportExport,Mathematica:ImportExport,Matlab:ImportExport,Sage:ImportExport}. However, because of the explained difficulties, those functions are very limited. In fact, they only work for expressions with unique semantic information, but fail for more complex expressions, such as our Jacobi polynomial example.

The \gls{nist} has therefore developed a set of semantic \LaTeX{} macros~\cite{DLMF:Macros}. Each of these macros is tied with a unique well-defined mathematical object and linked to the corresponding definition in the \gls{dlmf}~\cite{NIST:DLMF,NIST:DLMF:Paper,NIST:Handbook}. The semantic macros are used to semantically enhance the \gls{dlmf}. An outgrowth of the \gls{dlmf} project is the \gls{drmf}~\cite{DRMF:14,DRMF:15}, which defines more semantic macros, especially for \gls{opsf}. In table~\ref{tab:JacobiP-usecase} we have already seen an example of a semantic macro. We call \LaTeX{} expressions with semantic macros \textit{semantic \LaTeX{}} and without \textit{generic \LaTeX}, respectively. With semantic \LaTeX, it is possible to provide translations to \gls{cas}.

@@

Furthermore, note that the translator we will present in this thesis is based on not published code from the \gls{mlp} project and the \gls{pom}-tagger. We will also use the mentioned set of \Macro s, which are not published yet as well. Therefore, the current version of the translator is not online available yet. However, we are planning to publish the translator soon on the \gls{drmf} website\footnote{\url{http://drmf.wmflabs.org}}.

@@

\section{Goals of the project}
In this thesis, we will present an automatic tool for translations between semantic \LaTeX{} and the \gls{cas} \Maple{} and \Mathematica. The translation tool should be lite, in a sense that a user does not need a whole software package to use the program. It is also a part of the \gls{drmf} project and should provide interactive translations in the \gls{drmf}. Furthermore, it should provide additional information about the translation process and present solutions if an appropriate translation is not possible.

@@

In chapter~\ref{ch:evaluation} we will discuss, how we can verify or validate a translated expression. The chapter presents some approaches, such as round trip tests (section~\ref{sec:round-trip}) and numerical tests (section~\ref{sec:numerical-tests}). Extracted formulae from the \gls{dlmf} and \gls{drmf} created a comprehensive test suite for these approaches. Section~\ref{sec:test-summary} gives a summary of the results.

>>>>

A typical workflow of a scientist who writes a scientific publication is to use a \gls*{wp} to write the publication and one or more \gls*{cas} for Verification, Analysis and Visualization, among other tasks. Especially in \gls*{stem}, \LaTeX\footnote{Note that technically \LaTeX{} is not a \gls*{wp} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the standard to write in \gls*{stem} and we only focus on mathematical writings in this thesis, we categorize \LaTeX{} as a \gls*{wp} as well.} has become the de facto standard for writing scientific publications over the last 30 years~\cites{LATEX:Standard}[559]{DigitalTypo}{Knuth}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider a Jacobi polynomial~\cite[(18.3), table 1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)).
\end{equation}
This formula is written in \LaTeX{} as
\begin{equation}\label{eq:P-tex}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation}

While \LaTeX{} focuses on displaying mathematics, \gls*{cas} focus on computations and user friendly syntax. Therefore, each system (such as \LaTeX{}, \Maple{} or \Mathematica) uses its own representation and syntax. Hence, a writer needs to constantly translate mathematical expressions from one representation to another and back again. Table~\ref{tab:JacobiP-usecase} shows four different representations for the same formula~(\ref{eq:P}).

@@

Such translations can become complex, if there are non-obvious differences between the used systems. For example, the \gls*{cas} could use a different order of the arguments, define a normalization of the function for better performances or even use different domains for the variables. Multivalued functions are particularly difficult~\cite{AISC:MultivaluedFunctions}. A \gls*{cas} usually defines so called \textit{branch cuts} to compute principal values of multivalued functions~\cite{Maple:Cuts}. Thereby, implementations of multivalued functions are discontinuous. In general, positioning branch cuts follow some conventions. They are mostly defined conveniently, such as a straight line. However, since a branch cut is a curve, it also can be defined as a spiral or it can be defined for a different position in the complex plane. The position of branch cuts can therefore varies from \gls*{cas} to \gls*{cas}~\cite{Branches:acot}. Figure~\ref{fig:acot-cut-compare} illustrates two examples of different branch cut positioning for the inverse trigonometric arccotangent function. While \Maple{} defines the branch cut at ${[-\infty\iunit, -\iunit]}$, ${[\iunit,\infty\iunit]}$ (figure~\ref{fig:acot-cut1}), \Mathematica{} defines the branch cut at ${[-\iunit, \iunit]}$ (figure~\ref{fig:acot-cut2}).

@@

Hence, a \gls*{cas} user needs to fully understand the properties and special definitions (such as the position of branch cuts) in the \gls*{cas} to avoid mistakes during a translation~\cite{Maple:Cuts}. In consequence, a manual translation process is not only laborious, but also prone to errors. Note that this general problem has been named to automatic \gls*{p2c} conversion~\cite{POM-Tagger}.

An automatic translation process is desirable. Providing translations for mathematical \LaTeX{} expressions is difficult, because the semantic information is absent. Semantics is the meaning of an expression. Consider the example of the Jacobi polynomial in~(\ref{eq:P}). A scientific reader might be able to conclude information like $P$ indicating the Jacobi polynomial and $n$ is being a non-negative integer. However, for a computer it is difficult to understand the expression~(\ref{eq:P-tex}). Since \LaTeX{} is the de facto standard for mathematical expressions, most \gls*{cas} provide import and export functions for \LaTeX{} expressions~\cite{Maple:ImportExport,Mathematica:ImportExport,Matlab:ImportExport,Sage:ImportExport}. However, because of the explained difficulties, those functions are very limited. In fact, they only work for expressions with unique semantic information, but fail for more complex expressions, such as our Jacobi polynomial example.

The \gls*{nist} has therefore developed a set of semantic \LaTeX{} macros~\cite{DLMF:Macros}. Each of these macros is tied with a unique well-defined mathematical object and linked to the corresponding definition in the \gls*{dlmf}~\cite{NIST:DLMF,NIST:DLMF:Paper,NIST:Handbook}. The semantic macros are used to semantically enhance the \gls*{dlmf}. An outgrowth of the \gls*{dlmf} project is the \gls*{drmf}~\cite{DRMF:14,DRMF:15}, which defines more semantic macros, especially for \gls*{opsf}. In table~\ref{tab:JacobiP-usecase} we have already seen an example of a semantic macro. We call \LaTeX{} expressions with semantic macros \textit{semantic \LaTeX{}} and without \textit{generic \LaTeX}, respectively. With semantic \LaTeX, it is possible to provide translations to \gls*{cas}.

@@

Furthermore, note that the translator we will present in this thesis is based on not published code from the \gls*{mlp} project and the \gls*{pom}-tagger. We will also use the mentioned set of \Macro s, which are not published yet as well. Therefore, the current version of the translator is not online available yet. However, we are planning to publish the translator soon on the \gls*{drmf} website\footnote{\url{http://drmf.wmflabs.org}}.

@@

\section{Goals of the project}
In this thesis, we will present an automatic tool for translations between semantic \LaTeX{} and the \gls*{cas} \Maple{} and \Mathematica. The translation tool should be lite, in a sense that a user does not need a whole software package to use the program. It is also a part of the \gls*{drmf} project and should provide interactive translations in the \gls*{drmf}. Furthermore, it should provide additional information about the translation process and present solutions if an appropriate translation is not possible.

@@

In chapter~\ref{ch:evaluation} we will discuss, how we can verify or validate a translated expression. The chapter presents some approaches, such as round trip tests (section~\ref{sec:round-trip}) and numerical tests (section~\ref{sec:numerical-tests}). Extracted formulae from the \gls*{dlmf} and \gls*{drmf} created a comprehensive test suite for these approaches. Section~\ref{sec:test-summary} gives a summary of the results.

### 2018-07-09 chapters/abstract.tex
Apply article format and create a clean version

\cleardoublepage
\begin{abstract}
\setcounter{page}{3}
\glsresetall
\thispagestyle{plain}
This Master's thesis presents the development of a translation tool between representations of semantically enriched mathematical formulae in a \gls{wp} and its corresponding representations in a \gls{cas}. The representatives for the \gls{cas} are \Maple{} and \Mathematica{}, while the representative \gls{wp} is \LaTeX.

Semantic information of a mathematical formula in \LaTeX{} is rather given in the context than in the formula itself. Extracting the information is complicated and sometimes even impossible. However, a translation to a \gls{cas} is only achievable, if the semantic of the formula is unique. The \gls{nist} in the U.S. has developed a set of semantic \LaTeX{} macros for all \glslink{opsf}{orthogonal polynomials and special functions (OPSF)} defined in the \gls{dlmf}. Using these macros can disambiguate formulae and provide access to the semantic information.

Even if the semantics of a representation of one formula is unique, it does not need to match the semantics in another representation of the formula. For example, there are may be differences in the domains or a function in a \gls{cas} is normalized for a better performance. An important concept for complex and multivalued functions are \textit{branch cuts}. However, the positions of these cuts are not consistently defined and can vary from system to system. Therefore, a \gls{cas} user needs to fully understand the properties and definitions of the used functions in the \gls{cas}. Hence, a manual translation from one system to another is not only laborious, but also prone to errors.

This thesis explains the realization of the translator, and discusses and suggests the approaches to the problems mentioned above.
\end{abstract}

\newpage
% back side of title should be empty -> empty page here
\mbox{}
\begin{otherlanguage}{ngerman}
\begin{abstract}
\setcounter{page}{5}
\glsresetall
\thispagestyle{plain}
Im Rahmen der Masterarbeit wurde ein Programm entwickelt, welches semantisch angereicherte mathematische Formeln von einer Darstellung in einem \glslink{wp}{Textverarbeitungssystem (WP)} wie \LaTeX{} zu einer entsprechenden Darstellung in einem \glslink{cas}{Computeralgebrasystem (CAS)} wie \Maple{} oder \Mathematica{} vor- und zurückübersetzen kann.

Semantische Informationen von mathematischen Formeln in \LaTeX{} ergeben sich vor allem aus dem Kontext der Formeln. Diese Informationen zu extrahieren ist jedoch schwierig und in einigen Fällen bisher sogar unmöglich. Eine direkte Übersetzung zu einem \acrshort{cas} ist jedoch nur möglich, wenn die semantischen Informationen eindeutig sind. Das \gls{nist} in den USA hat dafür eine Liste von \LaTeX{} Makros erstellt, welche semantische Informationen für jedes \glslink{opsf}{orthogonale Polynom und jede spezielle Funktion (OPSF)} in der \gls{dlmf} bereitstellen. Mit Hilfe dieser Makros können Doppeldeutigkeiten vermieden und der Zugang zu den semantischen Informationen erleichtert werden.

Selbst wenn die Semantik in einer Darstellung eindeutig ist, muss diese nicht mit der Semantik in einer anderen Darstellung übereinstimmen. Es kann Unterschiede in den Definitionsmengen geben oder eine Normalisierung definiert sein, um eine schnellere Berechnung zu ermöglichen. Bei mehrwertigen, komplexen Funktionen spielen sogenannte \textit{Branch-Cuts} eine wichtige Rolle für \acrshort{cas}. Die Position dieser \textit{Branch-Cuts} folgt zwar allgemeinen Konventionen, diese müssen aber nicht zwangsläufig eingehalten werden. Der Umgang mit einem \acrshort{cas} erfordert daher gute Kenntnisse über die speziellen Eigenschaften und Definitionen. Deshalb ist eine Übersetzung von einer Darstellung (z.B. \LaTeX) zu einer Anderen (z.B. \Maple) per Hand nicht nur mühselig, sondern auch fehleranfällig.

In der vorliegenden Arbeit wird die Umsetzung des Übersetzers erläutert und Lösungsansätze für die oben genannten Probleme werden erörtert.
\end{abstract}
\end{otherlanguage}
\setcounter{page}{6}
\newpage
\mbox{}
\newpage

>>>>

\begin{abstract}
\glsresetall
\noindent
This Master's thesis presents the development of a translation tool between representations of semantically enriched mathematical formulae in a \gls*{wp} and its corresponding representations in a \gls*{cas}. The representatives for the \gls*{cas} are \Maple{} and \Mathematica{}, while the representative \gls*{wp} is \LaTeX.
%
Semantic information of a mathematical formula in \LaTeX{} is rather given in the context than in the formula itself. Extracting the information is complicated and sometimes even impossible. However, a translation to a \gls*{cas} is only achievable, if the semantic of the formula is unique. The \gls*{nist} in the U.S. has developed a set of semantic \LaTeX{} macros for all \gls*{opsf} defined in the \gls*{dlmf}. Using these macros can disambiguate formulae and provide access to the semantic information.
%
Even if the semantics of a representation of one formula is unique, it does not need to match the semantics in another representation of the formula. For example, there are may be differences in the domains or a function in a \gls*{cas} is normalized for a better performance. An important concept for complex and multivalued functions are \textit{branch cuts}. However, the positions of these cuts are not consistently defined and can vary from system to system. Therefore, a \gls*{cas} user needs to fully understand the properties and definitions of the used functions in the \gls*{cas}. Hence, a manual translation from one system to another is not only laborious, but also prone to errors.
%
This thesis explains the realization of the translator, and discusses and suggests the approaches to the problems mentioned above.
\end{abstract}
